---
title: "Contrasts:  Pre-Planned and Simultaneous"
output: learnr::tutorial
runtime: shiny_prerendered
author: "Prof. Jessica"
date: '`r Sys.Date()`'


---

```{r setup, include=FALSE}
library(learnr)
library(multcomp)
library(DescTools) # install first
library(readxl)
wheat3 <- read_excel("wheat3.xlsx")

wheat3$Type.f = factor(wheat3$Type,levels = c("winter","spring","hybrid")) 
fit3 = aov(ProteinConc ~ Type.f, data=wheat3)	 # perform the ANOVA computations and store

n = 24; trt = 3
s = sigma(fit3); s              # or, get from above output

library(gmodels)                  # install first
contrast = c(0, 1, -1)      # must match level orderings!!!
fit.contrast(fit3, "Type.f", contrast)    # performs the “contrast”
results = fit.contrast(fit3, "Type.f", contrast) 

contrast.point = results[1]  # point estimate
contrast.SE = results[2]  # standard error of point estimate
contrast.point; contrast.SE; alpha = 0.05
alpha.E = 0.05  # experimentwise error rate
h = 3
alpha.C = alpha.E/h

library(DescTools) # install first
```

#### <span style="color:Blue">**Intellectual Property:**</span>  
This tutorial is the intellectual property of the instructors and may not be reproduced outside of this course.


## Wheat data
We will be working with the wheat data throughout this tutorial, walking through the lecture code (reference Lectures 5 and 6)

```{r data, eval=F}
library(readxl)
wheat3 <- read_excel("wheat3.xlsx")
```
Note:  this is *not* in exercise format; the commands run automatically.

## Working with levels of a factor 

**Participation points**:  Which level is listed first in the default ordering?

When we make a variable into a factor for use in ANOVA and follow-up analyses, the levels will default to an alphabetical or numeric ordering.  
```{r defaultordering, exercise=TRUE}
factor(wheat3$Type)  # this is the factor values with the default levels
levels(factor(wheat3$Type) )  # or just take a look at the levels
boxplot(ProteinConc~Type,data=wheat3)
```

Often, this will *not* coincide with our preferred ordering.  So, we can use the <font face="Courier">levels=</font> argument in the <font face="Courier">factor</font> function:

   
```{r specifiedordering, exercise=TRUE}
wheat3$Type.f = factor(wheat3$Type,levels = c("winter","spring","hybrid"))  # specify level ordering
levels(wheat3$Type.f)   # just take a look at the levels
boxplot(ProteinConc~Type.f,data=wheat3)
```

## Preliminaries (review ANOVA commands)

**Participation points**:  What is the total sample size *n*?  What is the number of replications *r*?

Preliminaries - we will run ANOVA, as we have done previously, as well as storing important information from the output:


```{r preliminaries}
# ANOVA
fit3 = aov(ProteinConc ~ Type.f, data=wheat3)	 # perform the ANOVA computations and store
summary(fit3)	              # summarize the output, including ANOVA table
print(fit3)

n = 24; trt = 3
s = sigma(fit3); s              # or, get from above output
```
Note:  this is *not* in exercise format; the commands run automatically.

## Pre-planned contrasts

**Participation points**:  What is the P-value for a test of $$H_1:  \mu_2 \ne \mu_3$$?

Now, we are ready to estimate the planned contrast:  $$\mu_2-\mu_3$$

Note that, for this contrast, in class we identified the contrast coefficients:  $$k_1 = 0$$, $$k_2 = 1$$, $$k_3 = -1$$, so we will list those off in a variable called <font face="Courier">contrast</font>.  We then use the ``glht()`` function to compute inferential summaries (point estimate, standard error, test statistic, and P-value) for that contrast.



```{r backgrounddefine2, echo=F}
wheat3$Type.f = factor(wheat3$Type,levels = c("winter","spring","hybrid")) 
```

Note that the output from ``results`` is a list, so we use functions to pick elements off and run test / make interval:
```{r contrastfit, exercise=T}
contrast = c(0, 1, -1)      # must match level orderings!!!

# fit contrast
library(multcomp)
results = glht(fit3, mcp(Type.f=contrast))

# computed statistics
summary(results)
# computed interval
confint(results)
```







## Simultaneous contrasts - Bonferroni 

We start with the necessary setup, using experimentwise error rate of 0.05.  Since we compute the Bonferroni intervals for each of three possible comparisons, we use *h*=3:  
```{r bonfsetup, exercise=T}
alpha.E = 0.05  # experimentwise error rate
# specifcally for three Bonferroni intervals
h = 3
alpha.C = alpha.E/h
1-(alpha.C/2)
```

The three comparisons look at comparing each treatment mean to the average for the other two treatments. 
$$\mu_1-\frac{\mu_2+\mu_3}{2}$$

```{r bonfcomp1, exercise=T}
contrast1 = c(1,-.5,-.5)      # must fill in for chosen contrast

results1 = glht(fit3, mcp(Type.f=contrast1))
summary(results1)
confint(results1, level=1-alpha.C)
```

$$\mu_2-\frac{\mu_1+\mu_3}{2}$$

```{r bonfcomp2, exercise=T}
contrast2 = c(-.5,1,-.5)      # must fill in for chosen contrast

results2 = glht(fit3, mcp(Type.f=contrast2))
summary(results2)
confint(results2, level=1-alpha.C)
```

$$\mu_3-\frac{\mu_1+\mu_2}{2}$$

```{r bonfcomp3, exercise=T}
contrast3 = c(-.5,-.5,1)      # must fill in for chosen contrast

results3 = glht(fit3, mcp(Type.f=contrast3))
summary(results3)
confint(results3, level=1-alpha.C)
```


## Simultaneous contrasts - Scheffe 

Next, we use Scheffe with an overall bound on the experimentwise error rate, covering all possible contrasts.  We again look at comparing each treatment mean to the average for the other two treatments. 


$$\mu_1-\frac{\mu_2+\mu_3}{2}$$

```{r scheffecomp1, exercise=T}
library(DescTools) # install first

contrast1 = c(1,-.5,-.5)      # must fill in for chosen contrast
ScheffeTest(fit3,"Type.f", contrasts=contrast1, conf.level = 1-alpha.E)
```

$$\mu_2-\frac{\mu_1+\mu_3}{2}$$

```{r scheffecomp2, exercise=T}
contrast2 = c(-.5,1,-.5)      # must fill in for chosen contrast
ScheffeTest(fit3,"Type.f", contrasts=contrast2, conf.level = 1-alpha.E)
```

$$\mu_3-\frac{\mu_1+\mu_2}{2}$$

```{r scheffecomp3, exercise=T}
contrast3 = c(-.5,-.5,1)      # must fill in for chosen contrast
ScheffeTest(fit3,"Type.f", contrasts=contrast3, conf.level = 1-alpha.E)
```

## Simultaneous contrasts - pairwise comparisons:

**Participation points**:  Report each 95% confidence interval for the pairwise comparison $$\mu_3-\mu_2$$ for each of the four simultaneous methods:  Bonferroni, Scheffe, Tukey's HSD, and Fisher's LSD.

All pairwise using Bonferroni:
```{r allpairbonf, exercise=T}
library(DescTools) # install first

PostHocTest(fit3, method = "bonferroni", conf.level = 1-alpha.E)
```

All pairwise using Scheffe:
```{r allpairscheffe, exercise=T}
PostHocTest(fit3, method =  "scheffe",conf.level = 1-alpha.E)
```

All pairwise using Tukey's HSD:
```{r allpairhsd, exercise=T}
PostHocTest(fit3, method = "hsd", conf.level = 1-alpha.E)
```

All pairwise using Fisher's LSD:
```{r allpairlsd, exercise=T}
PostHocTest(fit3, method = "lsd", conf.level = 1-alpha.E)
```

