---
title: "Intro to R - Lesson 1:  The Very Basics"
output: learnr::tutorial
runtime: shiny_prerendered
author: "Prof. Jessica (Mathematics) and Dr. Haffner (Geography)"
date: '`r Sys.Date()`'


---

```{r setup, include=FALSE}
library(learnr)
x = c(4:6,10:12)
A = matrix(c(1:4,9:12),ncol=2)

```

#### <span style="color:Blue">**Intellectual Property:**</span>  
This tutorial is the intellectual property of the instructors and may not be reproduced outside of this course.

## Example code line ##

This code computes the answer to one plus one, change it so it computes two plus two:

   * Run the code
   * Type a different addition
   * Try some other arithmetic

**Note**: You can run a single line (with cursor in-line) by typing Ctrl-Enter or Command-Enter.

```{r trytoadd, exercise=T}
1+1
8*7
```

## Vectors and arrays ##
### Entering / manipulating data in R ### 

Suppose we have a small dataset that can easily be entered by hand. We enter the data in the main dialog box using the general vector function c() and the matrix function matrix().

Let's try entering some data.  Anytime we enter data, it needs to be "stored".  We need to:

   * Name a location (variable, or matrix, or data frame) to store the data
   * Assign (using either <- or =) the values to the storage location
   * Recall the values by typing the storage name

```{r storage, exercise=TRUE}
# try typing y; what happens?
# y is not a defined variable

x = 7
x     # a single value can be stored in the location x

x = c(4:6,10:12)
x     # a "column vector", which is what the c() command denotes, stored in x

A = matrix(c(1:4,9:12),ncol=2)
A     # we can us the matrix function to store a two-dimensional array

# there is also ways to make larger-dimensional arrays, but these are rarely used
```



## Data subsets ##
### Entering / manipulating data in R ### 

  We can choose a subset of a variable or array by using brackets, [ ], following the variable name.

```{r subsets, exercise=TRUE}
#this is a comment
print(c("x-vector is", x))
x[1]  # first element of x
x[1:3]  # first three elements of x
x[c(2,4)]  # second and fourth elements of x

pizza = c(9, 12, 15)  #call a variable whatever you like
pizza
piz_za = c(16,-3)  # adding symbols makes a new variable
piz_za
Pizza  = c(1, 8, -5)  # so does capitalization, importantly and annoyingly

A  # matrix
A[1:2,1]  # first two rows and first column
A[1:2,][,1]  # ditto
A[c(2,4),]  # second and fourth rows
A[,1]  # first column
```

## Manipulating variables ##
### Entering / manipulating data in R ### 

Mathematical functions in R work “element-wise”, meaning they are applied to each element in the vector / array.

```{r constant, exercise=TRUE}
x 
3*x
x-3

A
3*A
A-3
```


## Using data frames ##
### Entering / manipulating data in R ### 

A **data frame** is a structure with:

   * rows representing observations; 
   * columns representing variables; and 
   * column names (**headers**) attached to the variables.
   
Referencing the variables:

   * Ideally, use the operator $ to "peel off" a variable; or
   * can "attach" the data frame.
   
```{r dataframe, exercise=TRUE}
data(iris)  # this is a "base" data frame / set in R
iris[c(1, 51,101),]
iris$Sepal.Length[1:5]  # recognizes values

# try typing Sepal.Length; what happens when you try to run?

attach(iris)
 Sepal.Length  # now connected
detach(iris)  # remember to detach when you are done with analysis, if you attach! 
```

##  Comparison operators ##
### Entering / manipulating data in R ### 

We can use a variety of logical checks and comparisons to work with data:

   * Less than (or equal to):  < or <=
   * Greater than (or equal to):  > or >=
   * Equal to:  ==
   * Not equal to !=

Each comparison results in a logical vector, that is, a vector of TRUE or FALSE values, produced element-wise.

```{r comparisons, exercise=TRUE}
data(iris)  # this is a "base" data frame / set in R
iris$Petal.Length
iris$Petal.Length > 1.5
iris$Petal.Length <= 1.5
iris$Petal.Length == 1.5
iris$Petal.Length != 1.5

#using the logical vector to subset
iris$Sepal.Length[iris$Petal.Length == 1.5]

```

##  Simple functions ##

Functions in R are identified by a word, followed by parantheses with arguments inside:

*function*(*argument1*, *argument2*, ...)

Many are intuitive, but others may need searching.  You may search for a topic by:

??"*topic key words*"

Then use 

help(*function*) 

for details on its usage.

```{r functions, exercise=TRUE}
??"histogram"  # search for how to make a histogram

hist(x = iris$Petal.Length)  # make a histogram of Petal Lengths, all other arguments defaults
hist(x = iris$Petal.Length, breaks = 8)  # make a histogram of Petal Lengths, specifying 8 groups
hist(iris$Petal.Length, 8)  # make a histogram of Petal Lengths, requires that arguments be in correct order
```


